Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-09-23 Scripts/Manuscript.Rmd
First discussion draft

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@SDG14; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @CheungClimateReadyFish]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg;@Pinsky:2018cb;].

In the early 1980's the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl] partitioning the distribution of several species, thus, creating what we known today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@FAO; @Munro]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence shows that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@EuropeShift]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song]. Further south, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted towards Chilean waters adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. The objective of the current paper is to explore two concepts that will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries [@Link:2010ei].

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. If a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S]. A thread point can be translated to the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@PalaciosAbrantes:E&S]. Many shared fisheries operate this way [@EuropeShift; @GoMA; @NOAA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@PalaciosAbrantes:E&S; @Sumaila:E&S]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive species further away from their historic range limits [@Alexa?]. Assessing the ToE of shared stocks is as important as the change itself, as it could reduce potential international conflicts [@Spijkers:2017ij] and aid the strengthen of marine governance [@Pinsky:2018cb;  @Pinsky:2014by; @Link:2010ei].

@@

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@Palacios-Abrantes:xxxx]. Each species in within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM). The DBEM algorithm includes species physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members (See Estimating the Time of Emergence). 
## Changes in Stocks Share Ratio

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@Palacios-Abrantes:E&S]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

@@

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated the change in SSR (larger than 1 s.d) of transboundary species shared by neighboring EEZs by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). Most EEZs will average a change lower than 25% which is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). However, this result masks larger changes in SSR for specific species. For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. By early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR >< 1 s.d. of historical distribution will be relatively large (global mean of 69% of species in an EEZ). For 17 EEZs, 100% of their analyzed transboundary stocks will shift, while and only six EEZs will have less than 25% shifting stocks (Fig. 1A-Polygons). Regions where EEZs are relatively small and have multiple borders (e.g. The Caribbean) will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^{st}$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to the early $21^{st}$ century but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

<!-- # Discussion -->

<!-- The combination of the time of emergence and the revenue participation of the species could shed a light in the importance of the time of emergence in terms of management. For example, we identified that in Peru 12 species will emerge by 2038 in average, representing 92% of the country's fishing value. Moreover, our results show that Peruvean anchoveta, the second largest fishery of the world [@SOFIA] and the main fishery in Peru [@REF] have already emerged (2010), with its stock share ratio benefiting Chile (Fig. 1). -->

<!-- Form all significantly different variables, only latitudinal gradient postpone the ToE suggesting that the ToE will happen first at equatorial latitudes and later towards the poles. -->

<!-- Historically, environmentally-driven shits in shared stocks have cause problems and threaten stock's sustainability. **In early 1980's Canada and the US blablabla.** Many fish stocks managed by the European Union ICES regultaroy areas have been shifting since the 1970's [@EuropeShift]. Such shifts have already result in confrontations between nations as evidenced by the expansion of the Atlantic mackerel (*Scomber scombrus*) distribution reaching the EEZ of Iceland [@Spijkers:2017ij].  More recently, Chile and Peru have declare a cooperation strategy to mange the transboundary stock of Peruvean Anchovera [@Peru]. Furthermore, as we continue to enter the $21^{st}$ century, commercial marine species are expected to continue shifting (mostly poleward) potentilly compromising shared stocks management. By the end of the century, all EEZs of the world are proyected to have at least one new transboundary species with some of them having over 10 new stocks [@Pinsky]. Consequently, as species occupy new waters they leave historic areas and so, xx% of species are expected to exit EEZs by the end of che century [@Oremus]. This entering and exicting national jurisdicions will complicate fisheries management potentially creating confllict between neighboring nations [@Pinksy @Oremus:2020id]. -->

<!-- Conclusion -->

<!-- Preparing policy to species on the move is key to achieve sustainable development goals and proper governance of the world fisheries [@Pinksy]. -->

# Recycling text

Ignoring anthropocentric effects, a species biomass distribution follows a variation in time and space according to environmental preferences, species interactions and landscape features [@Huch]. Thus, a neighboring nation will not necessarily have, at all times  (e.g. months or years), a fixed SSR but rather a series of SSRs [@GoMA].

>>>>

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@UnitedNations2018; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @Cheung:2018dh]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

In the early 1980's the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl] partitioning the distribution of several species, thus, creating what we known today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@Munro:2004th]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence shows that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@Baudron:2020dj]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song:2017va]. Further south, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted towards Chilean waters adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. The objective of the current paper is to explore two concepts that will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries [@Link:2010ei].

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. If a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@OCSumaila]. A thread point can be translated to the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@OCPalacios]. Many shared fisheries operate this way having quota allocated based on some historical level [@Baudron:2020dj; @GoMA; @NOAA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@OCPalacios; @OCSumaila]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive species further away from their historic range limits. Assessing the ToE of shared stocks is as important as the change itself, as it could reduce potential international conflicts [@Spijkers:2017ij] and aid the strengthen of marine governance [@Pinsky:2018cb;  @Pinsky:2014by; @Link:2010ei; @Fogarty:2017eo].

@@

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@FishForVisa]. Each species in within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @FishForVisa]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region according to the UN and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM). The DBEM algorithm includes species physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members (See Estimating the Time of Emergence).

## Changes in Stocks Share Ratio

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@OCPalacios]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus, representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

@@

## Vulnerability index

We created a vulnerability index of fisheries management to shifts in transboundary species. The index is based on two factors; the importance of the shifting species to a country's total catch and revenue from transboundary species, and the time of emergence;

$$ index' = (P_c + P_v)/ToE $$
where *P_c* and *P_v* are the proportion that the shifting species' will have in the total catch and value generated from transboundary fisheries within a county's EEZs. Here, total catch and value are represented as the average of 2005 to 2014 according to the *Sea Around Us*. The index was then normalized to a 0 to 1 scale going from less vulnerable to more vulnerable:

$$ index = \frac{(index' - N_{min})}{N_{max}-N_{min}} $$

The premise of the index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Similarly, the index is capable to capture different sources of societal benefits (e.g. food security trough catch or livelihood trough value). We assume that the management of fisheries that provide large societal benefits will have more pressure from stakeholders while earlier times of emergence will result in less time to act, thus resulting in higher vulnerability.

## Caveats and uncertainties

We have identified three main sources of uncertainty in our analysis and point out ways to move forward in order to produce more reliable results to inform policy dealing with international fisheries under climate change. First, in this analysis we relay in one Earth System Model (GFDL-ESM2M) and climate change scenario (RCP 8.5). While the use of multiple ESMs would allow to capture the structural uncertainty of the models [@Frolicher:2016dg], using the GFDL-ESM2M results in conservative estimates as its projections are the less extreme from the CIMIP5 models [@Bopp:2013fg]. Addressing the uncertainty in future projections is key to advance marine governance for species on the move [@Pinsky:2018cb], thus, future efforts should look at the ensemble members of multiple ESM to capture the uncertainty embedded in the models. Secondly, we set an arbitrary threshold of $\pm$ 1 s.d. (67% confidence) to determine when a SSR and ToE would happen (see methods). We tested the sensitivity of our results to a larger threshold (e.g. $\pm$ 2 s.d for a 95% confidence). While the selected threshold is more relaxed than a 2 s.d., the distribution of results do not dramatically change between thresholds with a difference in ToE of about 4 years suggesting that this is a robust threshold (Fig. S4). Thirdly, in here we present the result based on 633 species that were previously identified as shared by multiple neighboring nations [@FishForVisa]. Thus, our analysis is carried at the species level, rather than a more detailed organizational level (e.g. fish stock) due to limited spatial-specific data on species sub populations (stocks). Reproducing our analysis with spatially-explicit stock data would allow to produce more certain results in changes in both SSR and ToE of transboundary stocks, and potentially identity different types of shared stock shifts [@Link:2010ei]. However, many regions of the world lack these detailed data. Moreover, in many EEZs of the world fisheries are often managed at the species level [@FishForVisa].

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@REF], *cowplot* [@REF], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated the change in SSR (with a 63% of confidence) of transboundary species shared by neighboring EEZs by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005) under a high emission climate change scenario. Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). Most EEZs will average a change lower than 25% which is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). However, this result masks larger changes in SSR for specific species. For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. By early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR >< 1 s.d. of historical distribution will be relatively large (global mean of 69% of species in an EEZ). For 17 EEZs, 100% of their analyzed transboundary stocks will shift, while and only six EEZs will have less than 25% shifting stocks (Fig. 1A-Polygons). Regions where EEZs are relatively small and have multiple borders (e.g. The Caribbean) will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^{st}$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to the early $21^{st}$ century but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

# Discussion

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings here are noteworthy in several ways. First, we identified that the historical share proportion of transboundary species will change by as soon as 2030 in most neighboring EEZs (Fig. 1). Second, we show that the current distribution of most transboundary species have already emerged from the historical record and that virtually all EEZs have seen at least one emerging species since 2006 (Fig. 2). Thirdly, we identify regional patterns of complex changes in SSR and levels of vulnerability of transboundary fisheries management to climate change (Fig. 3).

By 2030, changes in SSR will be seen across the globe, but with no clear pattern in the direction of change (Fig. 1). Poleward shifts in SSR will be seen in regions of the world with relative large EEZs and few boundaries like the North East Pacific. However, despite a general recognition that climate change is shifting species distributions poleward [@Cheung:2010dt; @Poloczanska:2016kk; @IPCC:2019tn], the Eastern Tropical Pacific will see an equatorial shift with EEZs from Mexico to Ecuador loosing SSR to their southern neighbors. This shift is consistent with regional modeling studies suggesting that by mid $21^{st}$ century marine species will displace between 50 and 100 km southward of Mexico's EEZ towards Peru under a high emission climate change scenario [@TayClarkeChp2]. Moreover, regions with presence of multiple frontiers like the Caribbean and the North East Atlantic will see a complex network of changes in SSR with no marked direction (Fig. 1B). These regions are known for being highly connected by both fish adult migration [@REF] and larvae dispersal [@Ramesh:2019va] leading to joint management plans [@Baudron:2020dj] and the call for new implemented shared policies [@Levin:2018in]. However, game theory suggest that the more number of parties negotiating, the harder it is for parties to reach an agreement [@Sumaila:2013], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting shared stocks.

We set out to evaluate the emergence characteristics of transboundary stocks across the global EEZs. Our results suggest that most of the EEZs of the world and more than half of the transboundary species already emerged since the early 2000s (Fig. 2). This is in alignment with previous research showing changes in the composition of marine catches since 1970's as evidence of climate change [@Cheung:2013gk]. Indeed, an expansion in the distribution of multiple marine species has been documented in several regions of the world [@Poloczanska:2016kk]. For example, in the early 2000s Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range reaching Washington waters where a new fishery rapidly developed [@Pinsky:2014by; @@Zeidberg:2007gs]. In the North East Atlantic, the emergence of Atlantic mackerel (*Scomber scombrus*) in Icelandic waters in 2010, lead to years of unsettled international conflict between multiple fishing nations [@Spijkers:2017ij]. The large amount of transboundary species emerging since early 2000s identified in this study can be partially attributed to the time of emergence of several environmental variables (e.g. temperature, oxygen, $CO_2$) that influence the distribution of marine species [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Mahlstein:2011fy]. It has been estimated that since the early 2000s, SST and oxygen ($O_2$) have emerged in 79% and 22.6% of the global ocean, respectively, and by 2080 SST will have emerged in 90% of the global ocean while $O_2$ will have doubled [@Rodggers:2015j].

Regional differences in ToE are likely the result of environmental factors, species vulnerability, and the characteristics of political boundaries. For example, the high number of stocks emerging early in the tropics suggest that tropical species are already on the move. The tropics are among the most species-rich biome [@Reygondeau:2019no] and have potentially experienced some of the largest declines in $O_2$ and greatest levels of warming since 2006 [@IPCC:2019tn; @Roemmich:2015id]. Here, marine species live close to the thermal tolerance making them highly vulnerable to warming waters [@IPCC:2014bz] consequently having to move towards deeper or sub-tropical waters [@Poloczanska:2016kk]. The premise of a shift already happening is supported by observational evidence of tropical species apearing in sub-tropical biomes [@Fogarty:2017eo] and catch-data analysis showing a reduction in sub-tropical species catches in tropical waters between 1970 and 2006 [@Cheung:2013gk]. In addition to species natural traits and the environmental pressure from climate change, the tropics comprise areas with relative small EEZs and many borders that match peaks of emerging stocks. Specifically, it seems like South East Asia, The Caribbean, and the Mediterranean are responsible for substantial increases in emerging stocks (Fig. 3).

An early ToE and large fisheries benefits could indicate high vulnerability of transboundary fisheries management to climate change (Fig. 3). For example, Polynesia and Micronesia are the only regions where the average ToE across species and EEZs already happened; 2019 and 2021, respectively (Fig. 5). Countries in these regions are highly dependent of pelagic-oceanic (e.g. tunas) and reef-associated transboundary species (e.g. *Speceies*) for both food security [@REF] and profits [@REF]. For example, in the Marshall Islands 16 transboundary stocks responsible for 92% and 94% of the transboundary fisheries revenue and catch, respectively, have, in averaged, emerged since 2018 resulting in high vulnerability index (Fig. 5). In this cases policy has to work against the clock and under potentially high societal pressure to address an issue that will impact important fisheries. However, a low vulnerability index does not necessarily translate to no-risk of international conflict or threat to stocks sustainability. While it gives countries more time to prepare for substantial changes, species with relatively small overall participation can still be a source of conflict as portraited by the Atlantic mackerel case [@FishForVisa; @Spijkers:2017ij].

# Conclusion

Preparing policy to species on the move is key to achieve sustainable development goals and proper governance of the world fisheries [@Pinsky:2018cb]. Here, we have addressed two fundamental steps to achieve sustainable international fisheries under a changing world [@Link:2010ei]; the degree of change that share species will have and the year in witch species will shift. Such results have direct implications for ocean governance. By identifying the degree location and time by which transboundary species will change from historical normal we provide critical information for regional management bodies and countries to anticipate the potential complications that climate change will bring to international fisheries management. While our results are not free of uncertainty and would benefit from more detailed analysis at localize levels, it represents a crucial step towards preparing ocean governance for species on the move. To address the issues raised in this paper, international fisheries management will require coordination between nations to generate adaptive, collaborative and equitable management strategies to address the uncertainties of a changing world.

# Supplements

```{r fig_s1, eval = F,  echo = F, fig.width=9, fig.height=12, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)"}

img <- readPNG("../Figures/Supp1.png")
grid.raster(img)

![Figure S1. World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points).](../Figures/Supp1.png)
```{r fig_s2, eval = F,  echo = F, fig.width=9, fig.height=12, fig.cap =  " Distribution of Stock Share Ratio percentage change by the early (2020-2040) and middle (2040-2060) 21st centurury relative to today (1970-2005)"}

img <- readPNG("../Figures/Supp2.png")
grid.raster(img)

![Figure S2. Distribution of Stock Share Ratio percentage change by the early (2021-2040) and middle (2040-2060) 21st centurury relative to today (1951-2005))](../Figures/Supp2.png)

```{r fig_s3, eval = F,  echo = F, fig.width=9, fig.height=12, fig.cap =  " Distribution of Stock Share Ratio percentage change by the early (2020-2040) and middle (2040-2060) 21st centurury relative to today (1951-2005)"}
